\mythesistitleNOTSYNPOSIS

### 2012-02-27 src/030chap3.tex
minor changes in TPL arbit trees -- more text around the figure. see diff.

\begin{lemma} 
  \label{lem:rooticppa}%
  Let $M$ be a matrix, $X$ be the directed graph whose vertices are in
  correspondence with the prime submatrices of $M$ and
  $\{\bP_1,\ldots,\bP_r\}$ be the partition of $X$ into in-trees by
  $\primeless$ defined earlier.  Let $T$ be the target tree and let
  $\{T_1, \ldots, T_r\}$ be a given partition of $T$ into vertex
  disjoint subtrees.  Then, for each $1 \leq i \leq r$, the set of
  prime submatrices $\bP_i$ has an ICPPL in $T_i$ if and only if the
  prime submatrix $mub(\bP_i)$ has an ICPPL on $T_i$ and all other
  matrices in $\bP_i$ have an ICPIA.
\end{lemma}
\begin{proof}\thesisspacing
  The proof is based on the fact that $\preccurlyeq$ is a partial
  order and $X$ is a directed graph which is a disjoint union of
  in-trees.  Each edge in the in-tree is a containment relationship
  among the supports of the corresponding submatrices. Therefore, any
  ICPPL to a prime submatrix that is not the $mub$ is contained in a
  path assigned to the sets in the parent matrix.  This follows from
  Lemma~\ref{lem:containment} and since $M_x \primeless mub(\bP_i)$
  for any prime submatrix $M_x \in \bP_i$.  Consequently, any ICPPL to
  the prime submatrix that is not $mub(\bP_i)$ is an ICPIA, and any
  ICPIA can be used to construct an ICPPL to the matrices
  corresponding to nodes in $\bP_i$ provided $mub(\bP_i)$ has
  an ICPPL in $T_i$. 
\end{proof}
Lemma \ref{lem:subicppa} and Lemma \ref{lem:rooticppa} point out two
algorithmic challenges in finding an ICPPL for a given hypergraph
$\cF$ on a target tree $T$.  
\begin{enumerate}
\item Finding $X$ and its partition into in-trees
  $\{\bP_1,\ldots,\bP_r\}$, can be done in polynomial time.  On the
  other hand, as per Lemma~\ref{lem:subicppa} we need to partition $T$
  into vertex disjoint subtrees $\{T_1, \ldots, T_r\}$ such that for
  each $i \in [r]$, the prime submatrices in $\bP_i$ have an ICPPL in
  $T_i$ (to be accurate, the prime submatrix $mub(\bP_i)$ has an ICPPL
  in $T_i$ and the rest of the prime submatrices have an ICPIA as
  claimed in Lemma~\ref{lem:rooticppa}).  This seems to be a
  challenging step, and it must be remarked that this step is easy
  when $T$ itself is a path, as each individual $T_i$ would be
  sub-paths.
\item The second algorithmic challenge is identified by
  Lemma~\ref{lem:rooticppa} which is to compute an ICPPL from a given
  tree to the matrix associated with $mub(\bP_i)$.
\end{enumerate}

@@

  \caption[\figtabsize Partial order on prime submatrices -
  containment partition]{\figtabsize
    Partial order on prime submatrices. The table on top illustrates
    the structural properties of prime matrices and partial order
    $\primeless$. A hypergraph $\cF$ with FTPL has hyperedges $S_1, \ldots,
    S_{12}$ which are represented by the binary matrix $M$. Empty
    entries denote zeros.
%     The graph
%     below is the target tree with each set $S_i, 1 \le i \le 12$ shown
%     as tree paths and intervals. Note that the sets that do not belong
%     to the $mub$s $M_1$ and $M_4$ are shown as intervals.
  }
  \label{fig:tplpotheory-matrix}
\end{figure}

@@

Figure~\ref{fig:tplpotheory-matrix} and
Figure~\ref{fig:tplpotheory-graph} illustrate the containment
partitions of a binary matrix and the claim of
Theorem~\ref{lem:rooticppa}. In Figure~\ref{fig:tplpotheory-matrix},
the feasibility bijection is explicitly shown by labeling the rows
(which represents elements in hypergraph's universe) with the vertices
of the target tree $T$, $v_1, \ldots, v_{12}$.  $M$ decomposed into
prime submatrices $M_1, \ldots, M_6$ which is collectively represented
by $\cP$. It can be verified that they correspond to the overlap
components in $\bO(\cF)$. The containment partitions of $\cP$ are
$\bP_1, \bP_2$ as shown. Note that the empty cells of the matrix
represent 0s -- they are left empty to make the containment partitions
more obvious.  Figure~\ref{fig:tplpotheory-graph} shows the target
tree of this TPL. The rectangles depict the sets allocated to the
columns in the $mub$ of $M$'s prime matrices, namely $M_1$ and
$M_4$. The rest of the columns are shown as intervals illustrating
Theorem~\ref{lem:rooticppa}.

\begin{algorithm}[h]
  \caption{Algorithm to find an ICPPL for a matrix $M$ on tree $T$:
    $main\_ICPPL(M, T$)}
  \label{al:icppa-main}
  \begin{algorithmic}[\lndisplay]
    \STATE Identify the prime submatrices. This is done by
    constructing the strict overlap graph and identify connected
    components.  Each connected component yields a prime submatrix.%
    \STATE Construct the partial order $\preccurlyeq$ on the set of
    prime submatrices.%
    \STATE Construct the partition $\bP_1, \ldots, \bP_r$ of the prime
    submatrices induced by $\preccurlyeq$%
    \STATE For each $1 \leq i \leq r$, check if all matrices except
    $mub(\bP_i)$ has an ICPIA.  If a matrix does not have ICPIA exit
    with a negative answer.  To check for the existence of ICPIA, use
    the result in \cite{nsnrs09}.%
    \STATE \label{ln:findtreepartition} Find a partition of $T_1, \ldots, T_r$ such that matrices
    in $mub(\bP_i)$ has an ICPPL in $T_i$.  If not such partition
    exists, exit with negative answer.%

@@

Algorithm~\ref{al:icppa-main} consolidates all that we described above
and gives the procedure to solve \CFTPL. It may be noted that all operations other than in
line~\ref{ln:findtreepartition} and line~\ref{ln:findrooticppl} are
known to be polynomial time solvable.

>>>>

\begin{lemma} 
  \label{lem:rooticppa}%
  Let $M$ be a matrix, $X$ be the directed graph whose vertices are in
  correspondence with the prime submatrices of $M$ and
  $\{\bP_1,\ldots,\bP_r\}$ be the partition of $X$ into in-trees by
  $\primeless$ defined earlier.  Let $T$ be the target tree and let
  $\{T_1, \ldots, T_r\}$ be a given partition of $T$ into vertex
  disjoint subtrees.  Then, for each $1 \leq i \leq r$, the set of
  prime submatrices $\bP_i$ has an ICPPL in $T_i$ if and only if the
  prime submatrix $mub(\bP_i)$ has an ICPPL on $T_i$ and all other
  matrices in $\bP_i$ have an ICPIA.
\end{lemma}
\begin{proof}\thesisspacing
  The proof is based on the fact that $\preccurlyeq$ is a partial
  order and $X$ is a directed graph which is a disjoint union of
  in-trees.  Each edge in the in-tree is a containment relationship
  among the supports of the corresponding submatrices. Therefore, any
  ICPPL to a prime submatrix that is not the $mub$ is contained in a
  path assigned to the sets in the parent matrix.  This follows from
  Lemma~\ref{lem:containment} and since $M_x \primeless mub(\bP_i)$
  for any prime submatrix $M_x \in \bP_i$.  Consequently, any ICPPL to
  the prime submatrix that is not $mub(\bP_i)$ is an ICPIA, and any
  ICPIA can be used to construct an ICPPL to the matrices
  corresponding to nodes in $\bP_i$ provided $mub(\bP_i)$ has
  an ICPPL in $T_i$. 
\end{proof}
Lemma \ref{lem:subicppa} and Lemma \ref{lem:rooticppa} point out two
algorithmic challenges in finding an ICPPL for a given hypergraph
$\cF$ on a target tree $T$.  
\begin{enumerate}
\item \label{it:challenge1} Finding $X$ and its partition into in-trees
  $\{\bP_1,\ldots,\bP_r\}$, can be done in polynomial time.  On the
  other hand, as per Lemma~\ref{lem:subicppa} we need to partition $T$
  into vertex disjoint subtrees $\{T_1, \ldots, T_r\}$ such that for
  each $i \in [r]$, the prime submatrices in $\bP_i$ have an ICPPL in
  $T_i$ (to be accurate, the prime submatrix $mub(\bP_i)$ has an ICPPL
  in $T_i$ and the rest of the prime submatrices have an ICPIA as
  claimed in Lemma~\ref{lem:rooticppa}).  This seems to be a
  challenging step, and it must be remarked that this step is easy
  when $T$ itself is a path, as each individual $T_i$ would be
  sub-paths.
\item \label{it:challenge2} The second algorithmic challenge is identified by
  Lemma~\ref{lem:rooticppa} which is to compute an ICPPL from the
  chosen subtree $T_i$
  to the matrix associated with $mub(\bP_i)$.
\end{enumerate}

Figure~\ref{fig:tplpotheory-matrix} and
Figure~\ref{fig:tplpotheory-graph} illustrate the containment
partitions of a binary matrix and the claim of
Lemma~\ref{lem:rooticppa}. In Figure~\ref{fig:tplpotheory-matrix}, the
feasibility bijection is explicitly shown by labeling the rows (which
represents elements in hypergraph's universe) with the vertices of the
target tree $T$, $v_1, \ldots, v_{12}$.  $M$ decomposed into prime
submatrices $M_1, \ldots, M_6$ which is collectively represented by
$\bP$. It can be verified that they correspond to the overlap
components in $\bO(\cF)$. The containment partitions of $\bP$ are
$\bP_1, \bP_2$ as shown. Note that the empty cells of the matrix
represent 0s -- they are left empty to make the containment partitions
more obvious.  Figure~\ref{fig:tplpotheory-graph} shows the target
tree of this TPL. The rectangles depict the tree paths allocated to
the columns in the $mub$ of $M$'s prime matrices, namely $M_1$ and
$M_4$. The rest of the columns are allocated intervals illustrating
Lemma~\ref{lem:rooticppa}.

Note that in this example the matrix has been explicitly labeled with
the vertices of the tree which indicates the solution but arriving at
this solution poses the above mentioned challenges.
Challenge~\ref{it:challenge1} in this example can be seen as
follows. Figure~\ref{fig:tplpotheory-graph} shows that $\bP_1$ has
been allocated the top subtree and $\bP_2$ has been allocated the
bottom subtree. Computationally this is not an easy decision. There is
no obvious indication (as far as we know) that the tree must be
decomposed at edge $(v_3, v_{11})$. If this was given, then one can
make the decision based on the difference between the subtrees, namely
sizes of the containment parts and the subtrees and conclude that
$\bP_1$ must be allocated the subtree $\{v_1, \ldots, v_7\}$ and
$\bP_2$ to the rest. What if the subtrees were of the same sizes, say
vertex $v_{14}$ in $T$, the fourteenth row in $M$ and column $S_9$ in
$M$ were not present? One has to try to solve
challenge~\ref{it:challenge2} given above to conclude that the bottom
(top) subtree indeed does not give an ICPPL to $\bP_1$ ($\bP_2$) and
thus $\bP_1$ ($\bP_2$) but be allocated the top (bottom) subtree.
Solving challenge~\ref{it:challenge2} is not easy (as far as we know)
either. By brute force, one can try all possible path labeling and
check for ICPPL. If all of them fail, one must backtrack to
challenge~\ref{it:challenge1} and try another subtree for the
containment part $\bP_i$.

Algorithm~\ref{al:icppa-main} consolidates all that we described in
this section and gives the procedure to solve \CFTPL. It may be noted
that all operations other than in line~\ref{ln:findtreepartition} and
line~\ref{ln:findrooticppl} are known to be polynomial time solvable.

@@

  \caption[\figtabsize Partial order on prime submatrices -
  containment partition]{\figtabsize
    Partial order on prime submatrices. The table illustrates
    the structural properties of prime matrices and partial order
    $\primeless$. A hypergraph $\cF$ with FTPL has hyperedges $S_1, \ldots,
    S_{12}$ which are represented by the binary matrix $M$. Empty
    entries denote zeros.
%     The graph
%     below is the target tree with each set $S_i, 1 \le i \le 12$ shown
%     as tree paths and intervals. Note that the sets that do not belong
%     to the $mub$s $M_1$ and $M_4$ are shown as intervals.
  }
  \label{fig:tplpotheory-matrix}
\end{figure}

@@

\begin{algorithm}[htb]
  \caption{Algorithm to find an ICPPL for a matrix $M$ on tree $T$:
    $main\_ICPPL(M, T$)}
  \label{al:icppa-main}
  \begin{algorithmic}[\lndisplay]
    \STATE Identify the prime submatrices. This is done by
    constructing the strict overlap graph and identify connected
    components.  Each connected component yields a prime submatrix.%
    \STATE Construct the partial order $\preccurlyeq$ on the set of
    prime submatrices.%
    \STATE Construct the partition $\bP_1, \ldots, \bP_r$ of the prime
    submatrices induced by $\preccurlyeq$%
    \STATE For each $1 \leq i \leq r$, check if all matrices except
    $mub(\bP_i)$ has an ICPIA.  If a matrix does not have ICPIA exit
    with a negative answer.  To check for the existence of ICPIA, use
    the result in \cite{nsnrs09}.%
    \STATE \label{ln:findtreepartition} Find a partition of $T_1, \ldots, T_r$ such that matrices
    in $mub(\bP_i)$ has an ICPPL in $T_i$.  If not such partition
    exists, exit with negative answer.%

@@
